Estimating the number of clusters using Cross Validation

### 2015-06-18 cvclust.tex
finish k-means generalized consistency

\newcommand{\CV}{\operatorname{CV}}
\newcommand{\E}{\operatorname{E}}
\newcommand{\dataX}{\mathfrak{X}}
\newcommand{\Xtrain}{X_{\text{train}}}
\newcommand{\Ytrain}{Y_{\text{train}}}
\newcommand{\Xtest}{X_{\text{test}}}
\newcommand{\Ytest}{Y_{\text{test}}}
\newcommand{\hGX}{\hat G^{X}}
\newcommand{\hGY}{\hat G^{Y}}
\newcommand{\bpi}{\bar \pi}
\newcommand{\bmuX}{\bar \mu^{X}}
\newcommand{\bmuY}{\bar \mu^{Y}}
\newcommand{\sA}{\mathcal{A}}
\newcommand{\sC}{\mathcal{C}}
\newcommand{\sE}{\mathcal{E}}
\newcommand{\sN}{\mathcal{N}}
\newcommand{\sX}{\mathcal{X}}

@@

\begin{lemma}\label{lem:pollard-continuous}\label{lem:pollard-last}
If $\int \|x\|^2 P(x) < \infty$, then the map $A \mapsto W(A, P)$ is
continuous on $\sA_k$ with respect to the topology induced by the Hausdoff
metric $H(\cdot, \cdot)$, defined for compact subsets $A, B$ of $\mathbb{R}^d$
by $H(A,B) < \delta$ if and only if every point of $A$ is within Euclidean
distance $\delta$ of at least one point of $B$, and vice versa.
\end{lemma}
\begin{proof}
This is a slight extension of the argument given at the end of Pollard's
Section~4.  If $A, B \in \sA_k$ and $H(A, B) < \delta$, then for each $b \in
B$ there exists an $a(b) \in A$ such that $\|b - a(b)\| < \delta$.  Then
\begin{align*}
  W(A, P) - W(B, P)
    &= \int \min_{a \in A} \|x - a\|^2 - \min_{b \in B} \|x - b\|^2 P(dx) \\
    &\leq \int \max_{b \in B} \{ \|x - a(b)\|^2 - \|x - b\|^2 \} P(dx) \\
    &\leq \int \sum_{b \in B} \{ (\|x - b\| + \delta)^2 - \|x - b\|^2 \} P(dx) \\
    &\leq \int \sum_{b \in B}  ( 2 \delta \|x - b\| + \delta^2) P(dx) \\
    &\leq 2 \delta \{M^{1/2} + k \max_{b \in B} \|b\|\} + k \delta^2,
\end{align*}
where $M = \int \|x\|^2 P(dx)$.  Thus,
\[
  |W(A, P) - W(B, P)|
    \leq 2 \delta \{M^{1/2} + k \max_{a \in A} \|a\|\} + 3 k \delta^2.
\]
Hence if $A \in \sA_k$ is any fixed set, then for every $\varepsilon > 0$, we
can find a $\delta$ small enough so that if $B \in \sA_k$ and $H(A, B) <
\delta$, then $|W(A,P) - W(B,P)| < \epsilon$.  Thus, $W(\cdot, P)$ is
continuous at $A$.
\end{proof}

\begin{lemma}\label{lem:kmeans-existence}

If $\int \|x\|^2 P(dx) < \infty$ and the support of $P$ contains at least $k$
points, then $m_1(P) > m_2(P) > \dotsb > m_k(P)$.  Further, there exists a
finite $M$ such that $W(\cdot, P)$ attains its infimum over $\sA_k$ and each
minimizer $\bar A$ is contained in the set
\(
  \sE_k = \{ A \in \sA_k : A \subset B(5M) \},
\)
where $B(5M)$ denotes the closed ball of radius~$5M$.
\end{lemma}
\begin{proof}
This argument is adapted from the proof in Section~3 of Pollard's paper.  We
first find an $M$ large enough to ensure that every minimizer of $W(\cdot, P)$
must have at least one point in $B(M)$.  Next, we impose additional
constraints on $M$ to ensure that all minimizers of $W(\cdot, P)$ are
contained in $B(5M)$.

Pick~$r$ so that the ball~$K$ of radius~$r$ centered at the origin has
positive $P$~measure.  Take $M$ to be large enough such that
\(
  (M - r)^2 P(K) > \int \|x\|^2 P(dx).
\)
Let $\sC_k$ denote the class of sets $A$ with at most $k$ points and at least
one point in in $B(M)$.  Let $A_0$ be the singleton set containing the origin.
If no point of $A$ is in $B(M)$, then
\(
  W(A, P) \geq (M - r)^2 P(K) > W(A_0, P).
\)
Put concisely, if $A \in \sA_k \setminus \sC_k$, then
\(
  W(A, P) > W(A_0, P) \geq m_k(P).
\)

Next, we proceed by induction on $k$.  If $k = 1$, then there is a unique
minimizer of $W(\cdot, P)$ over $\sA_k$, the set containing the mean $\int x
P(dx)$.  For $k > 1$, take as given that $m_{k-1}(P) = W(\bar A_{k-1}, P)$ for
some set $\bar A_{k-1} \in \sE_{k-1}$.  We will show that there is an $M$
large enough so that $W(\cdot, P)$ attains its infimum over $\sA_k$ and all
minimizers lie in $\sE_k$.

@@

We have now established that $m_k(P) = \inf \{ W(A,P) : A \in \sE_k \}$,
and that $W(\cdot, P)$ does not obtain its infimum over $\sA_k$ outside of
$\sE_k$.  The final step of the proof is to show $W(\cdot, P)$ obtains its
infimum over $\sE_k$.  This follows from the extreme value theorem, since
$\sE_k$ is compact and, by Lemma~\ref{lem:pollard-continuous}, the map $A
\mapsto W(A,P)$ is continuous over $\sA_k$.

@@

\begin{lemma}
\label{lem:pollard-compact}
\label{lem:pollard-first}
Suppose that $\int \|x\|^2 P(dx) < \infty$.  If $A_n$ minimizes $W(\cdot,
P_n)$, then, there exists a finite $M$ such that as if $n$ is large enough,
then, almost surely, $A_n$ is in the class of sets
\(
  \sE_k = \{ A \subset B(5M) : \text{$A$ contains $k$ or fewer points} \},
\)
where $B(5M)$ denotes the Euclidean ball of radius $5M$ centered at the
origin.
\end{lemma}

\begin{proof}
If $P$ has finite support, then the result follows immediately.  For the
remainder of the proof, then, we will assume that $P$ has infinite support.

In Section~3 of his paper, Pollard proves the result, assuming that there
is a unique set of $j$ points minimizing $W(\cdot, P)$ for each value of
$j = 1, \dotsc, k$.  His proof does not use this assumption directly, only
its implication that for each $j$ in the range $2, \dotsc, k$, there exists
a set with $j$ points satisfying $W(A,P) < m_{j-1}(P)$.  If we can establish the
implication, then the rest of his proof will apply.

We will prove the implication by construction.  Take $j$ to be in the range
$2, \dotsc, k$.  For any $\varepsilon > 0$, there exists a set
$B_{\varepsilon}$ containing at most $j-1$ points such that
$W(B_{\varepsilon}, P) < m_{j-1}(P) + \varepsilon$.  We will add a single
point to $B_{\varepsilon}$ to get a set $A_{\varepsilon}$ with $j$ or fewer
points satisfying $W(A, P) < m_{j-1}(P)$.

If we can show that $m_{j+1}(P) > m_{j}(P)$ for $j = 1, \dotsc, k-1$, then
we will have established the conditions required for Pollard's proof of the
analogous result in Section 3 of his paper.

\begin{lemma}\label{lem:polalrd-uniform}
If $\int \|x\|^2 P(x) < \infty$, then
\(
  \sup_{A \in \sE_k} |W(A, P) - W(A,P_n)| \to 0\text{ a.s.}
\)
\end{lemma}
\begin{proof}
Given at the beginning of Pollard's Section~4.
\end{proof}

>>>>

\newcommand{\CV}{\operatorname{CV}}
\newcommand{\E}{\operatorname{E}}
\newcommand{\dataX}{\mathfrak{X}}
\newcommand{\Xtrain}{X_{\text{train}}}
\newcommand{\Ytrain}{Y_{\text{train}}}
\newcommand{\Xtest}{X_{\text{test}}}
\newcommand{\Ytest}{Y_{\text{test}}}
\newcommand{\hGX}{\hat G^{X}}
\newcommand{\hGY}{\hat G^{Y}}
\newcommand{\bpi}{\bar \pi}
\newcommand{\bmuX}{\bar \mu^{X}}
\newcommand{\bmuY}{\bar \mu^{Y}}
\newcommand{\sA}{\mathcal{A}}
\newcommand{\sbA}{\mathcal{\bar A}}
\newcommand{\sC}{\mathcal{C}}
\newcommand{\sE}{\mathcal{E}}
\newcommand{\sN}{\mathcal{N}}
\newcommand{\sX}{\mathcal{X}}

@@

\begin{lemma}[Continuity]\label{lem:pollard-continuous}\label{lem:pollard-last}
If $\int \|x\|^2 P(x) < \infty$, then the map $A \mapsto W(A, P)$ is
continuous on $\sA_k$ with respect to the topology induced by the Hausdoff
metric $H(\cdot, \cdot)$, defined for compact subsets $A, B$ of $\mathbb{R}^d$
by $H(A,B) < \delta$ if and only if every point of $A$ is within Euclidean
distance $\delta$ of at least one point of $B$, and vice versa.
\end{lemma}
\begin{proof}
This is a slight extension of the argument given at the end of Pollard's
Section~4.  If $A, B \in \sA_k$ and $H(A, B) < \delta$, then for each $b \in
B$ there exists an $a(b) \in A$ such that $\|b - a(b)\| < \delta$.  Then
\begin{align*}
  W(A, P) - W(B, P)
    &= \int \min_{a \in A} \|x - a\|^2 - \min_{b \in B} \|x - b\|^2 P(dx) \\
    &\leq \int \max_{b \in B} \{ \|x - a(b)\|^2 - \|x - b\|^2 \} P(dx) \\
    &\leq \int \sum_{b \in B} \{ (\|x - b\| + \delta)^2 - \|x - b\|^2 \} P(dx) \\
    &\leq \int \sum_{b \in B}  ( 2 \delta \|x - b\| + \delta^2) P(dx) \\
    &\leq 2 \delta \{M^{1/2} + k \max_{b \in B} \|b\|\} + k \delta^2,
\end{align*}
where $M = \int \|x\|^2 P(dx)$.  Thus,
\[
  |W(A, P) - W(B, P)|
    \leq 2 \delta \{M^{1/2} + k \max_{a \in A} \|a\|\} + 3 k \delta^2.
\]
Hence if $A \in \sA_k$ is any fixed set, then for every $\varepsilon > 0$, we
can find a $\delta$ small enough so that if $B \in \sA_k$ and $H(A, B) <
\delta$, then $|W(A,P) - W(B,P)| < \epsilon$.  Thus, $W(\cdot, P)$ is
continuous at $A$.
\end{proof}

\begin{lemma}[Existence]\label{lem:kmeans-existence}
If $\int \|x\|^2 P(dx) < \infty$ and the support of $P$ contains at least $k$
points, then
\begin{enumerate}
  \item $m_1(P) > m_2(P) > \dotsb > m_k(P)$;
  \item $W(\cdot, P)$ achieves its infimum over $\sA_j$ for $j = 1, \dotsc,
    k$;
  \item the class of sets $\sbA_j = \{ A \in \sA_j : W(A,P) = m_{j}(P) \}$
    is compact (with respect to the Hausdorff metric on $\sA_j$) for $j = 1, \dotsc, k$.
\end{enumerate}
\end{lemma}
\begin{proof}
This argument is adapted from the proof in Section~3 of Pollard's paper.  We
first find an $M$ large enough to ensure that every minimizer of $W(\cdot, P)$
must have at least one point in $B(M)$.  Next, we impose additional
constraints on $M$ to ensure that all minimizers of $W(\cdot, P)$ are
contained in $B(5M)$.

Pick~$r$ so that the ball~$K$ of radius~$r$ centered at the origin has
positive $P$~measure.  Take $M$ to be large enough such that
\(
  (M - r)^2 P(K) > \int \|x\|^2 P(dx).
\)
Let $\sC_k$ denote the class of sets $A$ with at most $k$ points and at least
one point in in $B(M)$, the closed ball with radius $M$ centered at the origin.  Let
$A_0$ be the singleton set containing the origin.
If no point of $A$ is in $B(M)$, then
\(
  W(A, P) \geq (M - r)^2 P(K) > W(A_0, P).
\)
Put concisely, if $A \in \sA_k \setminus \sC_k$, then
\(
  W(A, P) > W(A_0, P) \geq m_k(P).
\)

Next, we proceed by induction on $k$.  If $k = 1$, then there is a unique
minimizer of $W(\cdot, P)$ over $\sA_k$, the set containing the mean $\int x
P(dx)$.  For $k > 1$, take as given that $m_{k-1}(P) = W(\bar A_{k-1}, P)$ for
some set $\bar A_{k-1} \in \sE_{k-1}$.  We will show that there is an $M$
large enough so that $W(\cdot, P)$ attains its infimum over $\sA_k$ and all
minimizers lie in $\sE_k$.  Since $A \mapsto W(\cdot, P)$ is continuous and
$\sE_k$ is compact, the class $\sbA_k$, the preimage of $\{ m_k(P) \}$ under this
map, is compact.

@@

We have now established that $m_k(P) = \inf \{ W(A,P) : A \in \sE_k \}$,
and that $W(\cdot, P)$ does not obtain its infimum over $\sA_k$ outside of
$\sE_k$.  The final step of the proof is to show $W(\cdot, P)$ obtains its
infimum over $\sE_k$.  This follows from the extreme value theorem, since
$\sE_k$ is compact and, by Lemma~\ref{lem:pollard-continuous}, the map $A
\mapsto W(A,P)$ is continuous over $\sA_k$.
\end{proof}

\begin{lemma}
\label{lem:pollard-compact}
Suppose that $\int \|x\|^2 P(dx) < \infty$, that the support of $P$ has at
least $k$ points, and that $k > 1$.  Let $A_n$ minimize $W(\cdot, P_n)$ over
$\sA_k$, and that $B_n$ minimize $W(\cdot P_n)$ over $\sA_{k-1}$.  If
$W(B_n, P) \to m_{k-1}(P)$ almost surely as $n \to \infty$, then
there exists a finite $M$ such that as if $n$ is large enough,
then, almost surely, $A_n$ is in the class of sets
\(
  \sE_k = \{ A \subset B(5M) : \text{$A$ contains $k$ or fewer points} \},
\)
where $B(5M)$ denotes the Euclidean ball of radius $5M$ centered at the
origin.
\end{lemma}

\begin{proof}
Pollard proves this result, in Section~3 of his paper,  assuming that there
is a unique set of $j$ points minimizing $W(\cdot, P)$ for each value of
$j = 1, \dotsc, k$.  His proof, which is very similar to the proof of
Lemma~\ref{lem:kmeans-existence}, does not use this assumption directly, only
its implication that $m_{k}(P) < m_{k-1}(P)$.  Since we have established his
result in Lemma~\ref{lem:kmeans-existence}, the remainder of his proof applies
without further modification.
\end{proof}

\begin{lemma}[Uniform SLLN]\label{lem:pollard-uniform}
If $\int \|x\|^2 P(x) < \infty$, then
\(
  \sup_{A \in \sE_k} |W(A, P) - W(A,P_n)| \to 0\text{ a.s.}
\)
\end{lemma}
\begin{proof}
Pollard proves this result at the beginning of his Section~4.
\end{proof}

\begin{theorem}[$k$-Means Consistency]
Suppose that $\int \|x\|^2 P(dx) < \infty$ and that $A_n$ minimizes
$W(\cdot, P_n)$ over $\sA_k$.  Then for all $n$ there exists an
$\bar A_n \in \sbA_k$ such that $H(A_n, \bar A_n) \to 0\text{ a.s.}$,
and $W(A_n, P_n) \to m_k(P)\text{ a.s.}$
\end{theorem}
\begin{proof}
The proof proceeds inductively over $k$.  For $k=1$, the result follows from
the strong law of large numbers.  For $k > 1$, assume that the result holds
for smaller values of $k$.  If the support of $P$ has fewer than $k$ points,
then the result is trivially true.  Otherwise, the induction hypothesis
implies that there exists a $B_n \in \sA_{k-1}$ with $W(B_n, P_n) \to
m_{k-1}(P)\text{ a.s.}$  Hence, Lemma~\ref{lem:pollard-compact} ensures that
there is a finite $M$ such that $A_n \in \sE_k$ almost surely for large
enough~$n$.  Take $M$ to be large enough so that $\sE_k$ also contains $\sbA_k$.

The set $\sE_k$ is compact and the map $f: \sE_k \to \mathbb{R}$ with $f(A) =
W(A, P)$ is continuous.  Thus, for every $\varepsilon > 0$ there exists an
$\eta > 0$ depending on $\varepsilon$ such that if $A \in \sE_k$ and $H(A,
\bar A) \geq \varepsilon$ for every $\bar A \in \sbA_k$, then $W(A,P) \geq
m_{k}(P) + \eta$.  Let $\bar A_0$ be any member of $\sbA_k$.  By construction,
$W(A_n, P_n) \leq W(\bar A, P_n)$.  Lemma~\ref{lem:pollard-uniform} ensures
that $W(\bar A_0, P_n) \to m_k(P)\text{ a.s.}$  Thus, for $n$ large enough,
$W(A_n, P_n) < m_{k}(P) + \eta$, which implies that there exists an $\bar A_n
\in \sbA_k$ with $H(A_n, \bar A_n) < \varepsilon$.

@@
